Kernel Square-Loss Exemplar Machines

### 2016-11-13 sec_SLEM.tex
Update sec_SLEM.tex

\section{Square loss exemplar machine}\label{lsesvm}
In this section, we revisit the exemplar SVM model presented by \cite{Efros11} as a more general exemplar classifier model. Then, we introduce the square loss exemplar machine (SLEM) by solving a simple variant this model and study its properties.
\subsection{Exemplar classifiers} \label{esvm}
We are given base features in $\RR^d$ at training time, one positive example $x_0$ in $\RR^d$ and a set of negative examples $X = [x_1, x_2,...,x_n]$ in $\RR^{d\times n}$, each column of $X$ representing one example by a vector in $\RR^d$. 
We are also given a loss function $l:\{-1, 1\}\times \RR\rightarrow\RR^+$. Learning an exemplar classifier from these examples amounts to minimizing the function 
\begin{equation}
J(\omega, \nu) = \theta \ l(1, \omega^Tx_0+\nu) +\dfrac{1}{n}\sumi l(-1, \omega^T x_i+\nu)+\dfrac{\lambda}{2}\|\omega\|^2, 
\label{eq:first}
\end{equation}
w.r.t. $\omega$ in $\RR^d$ and $\nu$ in $\RR$. In Equation (\ref{eq:first}), $\lambda$ and $\theta$ are respectively a regularization parameter on $\omega$ and a positive scalar adjusting the weight of the positive exemplar.
%\footnote{We could also acknowledge a parameter $\theta$ other than $\frac{1}{n}$ as a regularization to the error of each negative example. But this parameters seems to be less important to cross-validate. Also, setting $\theta=\frac{1}{n}$ simplify Equation (\ref{omega:solution}) and allows the use of Woodbury identity.}

The exemplar classifiers of $x_0$ with respect to $X$ are the weights $\omega^\star(x_0,X)$ that minimizes the loss function $J$:
\begin{equation}
\big(\omega^\star(x_0, X), \nu^\star(x_0, X)\big) = \underset{(\omega,\nu)\in\RR^d\times\RR}{\argmin} \ J(\omega, \nu). \label{omega:first}
\end{equation}
To shorten the notation, we refer to these solutions as $\omega^\star$ and $\nu^\star$\footnote{Depending on the loss function $l$, $\nu^\star(x_0,X)$ may be not unique.} when their arguments are implicit.

%\cite{Efros11} introduced ensembles of exemplar classifiers for object detection, but single exemplar classifiers have been used as replacement of the original feature $x_0$ for image classification and retrieval \cite{ZePe15} and detection \cite{GMPD12}. \RAF{Maybe this should be in the introduction instead.}
Applications of ESVM use the hinge loss function, which guarantees that $J$ is a convex function. The solution of Equation (\ref{omega:first}) can be thus found by stochastic gradient descent~\cite{bottou10} individually for each positive sample. The next subsection show how to calculate all exemplar classifiers simultaneously by changing the loss function.

\subsection{Square loss function}\label{slem:intro}
Now, let us study the same learning problem for the square-loss function $l(y,\hat{y}) = \frac{1}{2}(y-\hat{y})^2$. As in the case of the hinge loss, Equation (\ref{eq:first}) is a convex problem. 
However it is now a ridge regression problem, whose unique solution can be found in closed form as
\begin{align}
\begin{cases}
\vspace{3 mm}
\omega^\star &= \dfrac{2\theta}{\theta+1}U^{-1}(x_0-\mu), \\
%\vspace{3 mm}
\nu^\star &= \dfrac{\theta-1}{\theta+1}-\dfrac{1}{\theta+1}(\theta x_0+\mu)^T\omega^\star,
\end{cases}
\label{omega:solution}
\end{align}
%\begin{align}
%\nu^\star &= \dfrac{\theta-1}{\theta+1}-\dfrac{1}{\theta+1}(\theta x_0+\mu)^T\omega^\star, \label{eq:nustar}\\
%\omega^\star &= \dfrac{2\theta}{\theta+1}U^{-1}(x_0-\mu) \label{eq:wstar}
%\end{align}
where:
\begin{align}
\begin{cases}
\mu = &\frac{1}{n}\sum_{i=1}^n x_i,\\ \\
U = &\frac{1}{n}XX^T-\mu\mu^T\\
&+\frac{\theta}{\theta+1}(x_0-\mu)(x_0-\mu)^T+\lambda\mathrm{Id}_d. \label{eq:U}
\end{cases}
\end{align}
Equation (\ref{omega:solution}) shows how to solve (\ref{eq:first}) simultaneously for all exemplars.
%Replacing the hinge loss by the square loss offers a more compact solution, but not necessarily more efficient.

\textbf{Woodbury identity.} 
We can modify Equation \ref{omega:solution} by simplifing Equation \ref{eq:U}.
Let us define $A = \frac{1}{n}XX^T-\mu\mu^T +\lambda\mathrm{Id}_d$ our regularized covariance matrix and assume its inverse $A^{-1}$ known. 
%$A$ is the covariance matrix of the negative samples $X$\footnote{added of a small term proportional to the identity matrix to insure $A$ is positive-definite.}. Let us assume its inverse $A^{-1}$ is known. 
The matrix $U$ now reads $U = A + \frac{\theta}{\theta+1}\delta\delta^T$, where $\delta=x_0-\mu$ is the centralized (w.r.t. the negative mean) positive sample. The Woodbury identity \cite{woodbury} gives us
\begin{equation}
U^{-1} = A^{-1} -\dfrac{\theta}{\theta\delta^TA^{-1}\delta+ \theta+1}A^{-1}\delta^T\delta A^{-1}. \label{invU}
\end{equation}
Substituting (\ref{invU}) in (\ref{omega:solution}) yields
\begin{equation}
\begin{split}
\omega^\star &= \dfrac{2\theta}{\theta +1}\left(A^{-1}\delta - \dfrac{\theta}{\theta\delta^TA^{-1}\delta+ \theta+1} A^{-1}\delta (\delta^TA^{-1}\delta)\right)\\
&= \dfrac{2\theta}{\theta\delta^TA^{-1}\delta+ \theta+1} A^{-1}\delta.\label{Wood:omega}
\end{split}
\end{equation}

Note that the positive sample weight $\theta$ does not influence the direction of the optimal vector $\omega^\star$, only its norm. This means that if search and ranking are based on the normalized feature $\frac{1}{\|\omega^\star\|}\omega^\star$, $\theta$ does not influence the matching score of the SLEM vectors of two different images. This sets SLEM appart from ESVM that requires this parameter to be calibrated \cite{Efros11,ZePe15}. The normalized exemplar classifier $\omega^\star$ is therefore a linear transformation of $x_0-\mu$
%Indeed, if $\omega$ and $\omega'$ are the $d$-dimensional SLEM vectors of $x$ and $x'$, respectively, their cosine similarity reads:
%we can denote $s$ the matching score scalar function defined in $\RR^d\times \RR^d$, which is given by
%\begin{equation}
%s(\omega, \omega') = \dfrac{\omega^T \omega'}{\|\omega\| \|\omega'\|} = \frac{(x-\mu)^TA^{-2}(x'-\mu)}{\|A^{-1}(x-\mu)\|\|A^{-1}(x'-\mu)\|},\label{match:score}
%\end{equation}
%which does not depend on the value of $\theta$. 
%This means that the weight of the positive sample can be fixed at any positive value without changing matching scores.

@@

Let us now reanalyse Equation \ref{eq:first} for a particular case. Suppose that we have multiple positive samples. It can be shown that in this case, the corresponding linear classifier of Equation (\ref{eq:first}) for the square-loss is also given by
(\ref{omega:solution}), where $x_0$ denotes this time the center of mass
of the positive samples, {\em if} these samples have the {\em same} covariance matrix $\Sigma$ as the negative samples $X$.

@@

Note that SLEM is a generalized version of LDA. Indeed, when $\lambda=0$ in SLEM (i.e. no regularization), $A = \Sigma$, meaning $\omega_{LDA}$ of Eq. (\ref{eq:lda}) and $\omega^\star$ of Eq. (\ref{Wood:omega}) have the same direction.
%This observation has been equally made by \cite{Koba15}.

Many interesting properties of LDA has been used recently for classification tasks \cite{GMPD12,HMR12} and, more recently, for image retrieval with convolutional neural network activations \cite{babenko15}.
By simply generalizing LDA, we hope to obtain superior results.

>>>>

\section{Square loss exemplar machine}\label{lsesvm}
In this section, we revisit the exemplar SVM model presented by \cite{Efros11} as an instance of a more general family of classifiers. Then, we introduce the square loss exemplar machine (SLEM) by solving a simple variant this model and study its properties.
\subsection{Exemplar classifiers} \label{esvm}
We are given base features in $\RR^d$ at training time, one positive example $x_0$ in $\RR^d$ and a set of negative examples $X = [x_1, x_2,...,x_n]$ in $\RR^{d\times n}$, each column of $X$ representing one example by a vector in $\RR^d$. 
We are also given a loss function $l:\{-1, 1\}\times \RR\rightarrow\RR^+$. Learning an exemplar classifier from these examples amounts to minimizing the function 
\begin{equation}
J(\omega, \nu) = \theta \ l(1, \omega^Tx_0+\nu) +\dfrac{1}{n}\sumi l(-1, \omega^T x_i+\nu)+\dfrac{\lambda}{2}\|\omega\|^2, 
\label{eq:first}
\end{equation}
w.r.t. $\omega$ in $\RR^d$ and $\nu$ in $\RR$. In Eq. (\ref{eq:first}), $\lambda$ and $\theta$ are respectively a regularization parameter on $\omega$ and a positive scalar adjusting the weight of the positive exemplar.
%\footnote{We could also acknowledge a parameter $\theta$ other than $\frac{1}{n}$ as a regularization to the error of each negative example. But this parameters seems to be less important to cross-validate. Also, setting $\theta=\frac{1}{n}$ simplify Equation (\ref{omega:solution}) and allows the use of Woodbury identity.}

Given a cost $l$, we define the correspondent \emph{exemplar classifiers} of $x_0$ with respect to $X$ are the weights $\omega^\star(x_0,X)$ that minimizes the loss function $J$:
\begin{equation}
\big(\omega^\star, \nu^\star\big) = \underset{(\omega,\nu)\in\RR^d\times\RR}{\argmin} \ J(\omega, \nu). \footnote{Depending on the loss function $l$, $\nu^\star(x_0,X)$ may be not unique.}
%\big(\omega^\star(x_0, X), \nu^\star(x_0, X)\big) = \underset{(\omega,\nu)\in\RR^d\times\RR}{\argmin} \ J(\omega, \nu). 
\label{omega:first}
\end{equation}
%To shorten the notation, we refer to these solutions as $\omega^\star$ and $\nu^\star$ when their arguments are implicit.

%\cite{Efros11} introduced ensembles of exemplar classifiers for object detection, but single exemplar classifiers have been used as replacement of the original feature $x_0$ for image classification and retrieval \cite{ZePe15} and detection \cite{GMPD12}. \RAF{Maybe this should be in the introduction instead.}
The exemplar SVM~\cite{Efros11,Efros12} is an instance of this model where $l$ is the hinge-loss. 
%Applications of ESVM use the hinge loss function, which guarantees that $J$ is a convex function. 
The solution of Eq. (\ref{omega:first}) can be thus found by stochastic gradient descent~\cite{bottou10} individually for each positive sample. The next subsection show how to calculate all exemplar classifiers simultaneously by changing the loss function.

\subsection{Square loss function}\label{slem:intro}
Now, let us study the same learning problem for the square-loss function $l(y,\hat{y}) = \frac{1}{2}(y-\hat{y})^2$. As in the case of the hinge loss, Equation (\ref{eq:first}) is a convex problem. 
However it is now a ridge regression problem, whose unique solution can be found in closed form as
\begin{align}
\begin{cases}
\vspace{3 mm}
\omega^\star &= \dfrac{2\theta}{\theta+1}U^{-1}(x_0-\mu), \\
%\vspace{3 mm}
\nu^\star &= \dfrac{\theta-1}{\theta+1}-\dfrac{1}{\theta+1}(\theta x_0+\mu)^T\omega^\star,
\end{cases}
\label{omega:solution}
\end{align}
%\begin{align}
%\nu^\star &= \dfrac{\theta-1}{\theta+1}-\dfrac{1}{\theta+1}(\theta x_0+\mu)^T\omega^\star, \label{eq:nustar}\\
%\omega^\star &= \dfrac{2\theta}{\theta+1}U^{-1}(x_0-\mu) \label{eq:wstar}
%\end{align}
where:
\begin{align}
\begin{cases}
\mu = &\frac{1}{n}\sum_{i=1}^n x_i,\\ \\
U = &\frac{1}{n}XX^T-\mu\mu^T\\
&+\frac{\theta}{\theta+1}(x_0-\mu)(x_0-\mu)^T+\lambda\mathrm{Id}_d. \label{eq:U}
\end{cases}
\end{align}
Equation (\ref{omega:solution}) shows how to solve (\ref{eq:first}) simultaneously for all exemplars.
\cite{lssvm}
%Replacing the hinge loss by the square loss offers a more compact solution, but not necessarily more efficient.

\textbf{Woodbury identity.} 
We can simplify Eq. (\ref{omega:solution}) by modifying $U$ in Eq. (\ref{eq:U}).
Let us define $A = \frac{1}{n}XX^T-\mu\mu^T +\lambda\mathrm{Id}_d$ our regularized covariance matrix and assume its inverse $A^{-1}$ known. 
%$A$ is the covariance matrix of the negative samples $X$\footnote{added of a small term proportional to the identity matrix to insure $A$ is positive-definite.}. Let us assume its inverse $A^{-1}$ is known. 
The matrix $U$ now reads $U = A + \frac{\theta}{\theta+1}\delta\delta^T$, where $\delta=x_0-\mu$ is the centralized (w.r.t. the negatives' mean) positive sample. The Woodbury identity \cite{woodbury} gives us
\begin{equation}
U^{-1} = A^{-1} -\dfrac{\theta}{\theta\delta^TA^{-1}\delta+ \theta+1}A^{-1}\delta^T\delta A^{-1}. \label{invU}
\end{equation}
Substituting (\ref{invU}) in (\ref{omega:solution}) yields
\begin{equation}
\begin{split}
\omega^\star &= \dfrac{2\theta}{\theta +1}\left(A^{-1}\delta - \dfrac{\theta}{\theta\delta^TA^{-1}\delta+ \theta+1} A^{-1}\delta (\delta^TA^{-1}\delta)\right)\\
&= \dfrac{2\theta}{\theta\delta^TA^{-1}\delta+ \theta+1} A^{-1}\delta.\label{Wood:omega}
\end{split}
\end{equation}

Note that the positive sample weight $\theta$ does not influence the direction of the optimal vector $\omega^\star$, only its norm. This means that if search and ranking are based on the normalized feature $\frac{1}{\|\omega^\star\|}\omega^\star$, $\theta$ does not influence the matching score of the SLEM vectors of two different images. This sets SLEM appart from ESVM which requires this parameter to be calibrated \cite{Efros11,ZePe15}. We can thus set $\theta=\frac{1}{n}$ for the remaining of this work.
%The normalized exemplar classifier $\omega^\star$ is therefore a linear transformation of $x_0-\mu$.
%Indeed, if $\omega$ and $\omega'$ are the $d$-dimensional SLEM vectors of $x$ and $x'$, respectively, their cosine similarity reads:
%we can denote $s$ the matching score scalar function defined in $\RR^d\times \RR^d$, which is given by
%\begin{equation}
%s(\omega, \omega') = \dfrac{\omega^T \omega'}{\|\omega\| \|\omega'\|} = \frac{(x-\mu)^TA^{-2}(x'-\mu)}{\|A^{-1}(x-\mu)\|\|A^{-1}(x'-\mu)\|},\label{match:score}
%\end{equation}
%which does not depend on the value of $\theta$. 
%This means that the weight of the positive sample can be fixed at any positive value without changing matching scores.

@@

It is interesting to note the relationship between SLEM and the classical linear discriminant analisys (LDA).
Let us reanalyse Eq. (\ref{eq:first})) an suppose that we have multiple positive samples. It can be shown that in this case, the corresponding linear classifier of Eq. (\ref{eq:first}) for the square-loss is also given by
(\ref{omega:solution}), where $x_0$ denotes this time the center of mass
of the positive samples {\em if} these samples have the {\em same} covariance matrix $\Sigma$ as the negative samples $X$.

@@

This shows that SLEM is a generalized version of LDA: Indeed, when $\lambda=0$ in SLEM (i.e. no regularization), $A = \Sigma$, the vectors $\omega_{LDA}$ of Eq. (\ref{eq:lda}) and $\omega^\star$ of Eq. (\ref{Wood:omega}) have the same direction.
%This observation has been equally made by \cite{Koba15}.
Many interesting properties of LDA has been used recently for classification tasks \cite{GMPD12,HMR12} and, more recently, for image retrieval with convolutional neural network activations \cite{babenko15}.
With our simple generalization of LDA, we hope to obtain superior results.
